---
title: "Clean and merge the Job Prep Datasets"
author: "Halle Prine"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(here)

# Load custom functions
source(here("functions", "score_likert_scale.R"))
```

## Document Purpose

This document outlines the data cleaning process for the Job Preparation Study, including validation and formatting of IDs, cleaning of demographic, GPA, and survey data, and creation of composite variables needed for analysis.

## Data Import

First we will import the data and clean along the way whatever is necessary to clean first and foremost. 

```{r}
# Import raw demographics with unknown structure (extra column included)
demographics <- read_csv(here("Data", "Job Prep Demographics.csv"),
                         col_names = FALSE,
                         col_types = cols(.default = "c"))

# Import GPA Data
gpa <- read_csv(here("Data", "Job Prep GPA Data.csv"),
                skip = 5,
                col_types = "cnnfnnnn",
                col_names = c("ID", "school_code", "age", "gender",
                              "s1_credits", "s1_gpa", "s2_credits", "s2_gpa"))

# Import Survey Data
survey <- read_csv(here("Data", "Job Prep Survey Results.csv"),
                   skip = 4,
                   col_types = "cnnnnnnnnnnnnnn",
                   col_names = c("ID", paste0("item_", 1:14)))
```


## Data Cleaning
# Demographics Cleaning
Now we will clean each data frame starting with demographics 

```{r}
# Assign column names
colnames(demographics) <- c("ID", "school_code", "age", "gender", "discipline_raw")

# Flag and fill discipline rows
demographics <- demographics %>%
  mutate(discipline_marker = if_else(!is.na(ID) & is.na(school_code) & is.na(age) & is.na(gender), ID, NA_character_)) %>%
  fill(discipline_marker)

# Remove header junk and non-participant rows
demo_clean <- demographics %>%
  filter(!ID %in% c("ID", "Job Preparation Study", "Demographic Data"),
         !is.na(school_code), !is.na(age)) %>%
  transmute(ID = as.character(ID),
            school_code = as.character(school_code),
            age = as.numeric(age),
            gender = gender,
            discipline = discipline_marker)

# Remove identified junk rows and column
demo_tidy <- demographics %>%
  slice(-c(1:7, 146:150, 304:308, 464:468, 618:622)) %>%
  select(-discipline_raw)
```

# GPA Data Check and Cleanup

```{r}
# Summary check of GPA values
gpa %>%
  summarise(
    min_s1 = min(s1_gpa, na.rm = TRUE),
    max_s1 = max(s1_gpa, na.rm = TRUE),
    min_s2 = min(s2_gpa, na.rm = TRUE),
    max_s2 = max(s2_gpa, na.rm = TRUE)
  )

# Identify GPA values outside 1.0–4.3
gpa %>%
  filter(s1_gpa < 1 | s2_gpa < 1 | s1_gpa > 4.3 | s2_gpa > 4.3)

# Compute weighted GPA
gpa <- gpa %>%
  mutate(GPA_weighted = (s1_gpa * s1_credits + s2_gpa * s2_credits) / (s1_credits + s2_credits))
```

# Survey Data Cleaning

```{r}
# Our function that reverse codes and calculates composite scores
survey_tidy <- score_likert_scale(
  survey,
  scale_max = 6,
  reverse_items = c("item_4", "item_9", "item_12", "item_13")
) %>%
  rename(optimism_score = total_score)
```


## ID Checks

```{r}
# Check uniqueness and formatting of IDs in each dataset
demographics %>% count(ID) %>% filter(n > 1)  
gpa %>% count(ID) %>% filter(n > 1)
survey %>% count(ID) %>% filter(n > 1)

# Check format (e.g., all numeric or 3-digit strings)
all_ids_valid <- function(id_vector) {
  all(str_detect(id_vector, "^\\d{3}$"))  
}

all_ids_valid(demo_tidy$ID)
all_ids_valid(gpa$ID)
all_ids_valid(survey$ID)
```

## Data Merging 

Now we will merge the data into one analytical dataset. Inorder to retain all participants from the demographics dataset (demo_tidy) we decided to use a left join and just identify missing survey or GPA data
```{r}
# Make sure all IDs are character
demo_tidy <- demo_tidy %>% mutate(ID = as.character(ID))
gpa <- gpa %>% mutate(ID = as.character(ID))
survey_tidy <- survey_tidy %>% mutate(ID = as.character(ID))

# Merge datasets using left_join to retain all demographics
analytic_job_data <- demo_tidy %>%
  left_join(gpa, by = "ID") %>%
  left_join(survey_tidy, by = "ID")

# Check how many are missing GPA or survey data
summary(analytic_job_data)

# Count participants missing key fields
analytic_job_data %>%
  summarise(
    missing_gpa = sum(is.na(s1_gpa) | is.na(s2_gpa)),
    missing_survey = sum(is.na(optimism_score))
  )

# Due to the same columns names were were suffled and need to re-named
analytic_job_data <- analytic_job_data %>%
  rename(
    school_code = school_code.x,
    age = age.x,
    gender = gender.x
  )

# Additionally we need to take out repetitive columns 
analytic_job_data <- analytic_job_data %>%
  select(-c(6, 7, 8))


write_csv(analytic_job_data, "analytic_job_final.csv")
```
